Comment on Cumulative Excess Deaths in New Zealand in the COVID-19 Era: Biases from Ignoring Changes in Population Growth Rates

### 2024-07-07 rawfiles/letter_to_ed_docx.qmd
reviewer changes

This is a response to the [@gibson2024] paper in which, to give my summary of the paper, Gibson asserts that New Zealand authorities' measurement of mortality in the pandemic period is based on a flawed metric. Gibson's own metric, based on regression of deaths with an additive component of total population, is asserted as a more accurate model and produces approximately 4% excess mortality for the 2020-2022 period. If New Zealand authorities are using flawed metrics with significantly different results to reality, this would have profound governance and policy implications.

@@

For centuries, it has been established practice among actuaries, demographers, and health researchers to measure mortality using age-adjusted death rates where such data is available [@price1773]. This method supersedes the use of Crude Mortality Rates[^2] by acknowledging that different ages have different risks of death, so comparing age-based risk of death on a uniform basis. In New Zealand, tutorials on making mortality comparisons by using death rates by age and standardising to a reference population have been published since the 1890s [@didsbury1891].

@@

Age Standardised Mortality directly adjusts for both population growth and aging, as it is based on the number of deaths of the current resident population divided by the current resident population, stratified by age group and time period. This directly measures the population change effects Gibson is trying to model.

@@

The existence of centuries of mortality measurement establishes that it is best practice in mortality calculations to recognise that people of different ages die at different rates not mentioned in Gibson's article. While New Zealand medical professionals have used the Karlinsky and Kobak method/ Our World in Data graphs in media presentations, it is not established if that was for the designed purpose while still being aware of more accurate methods or because K & K was considered definitive. We can check if New Zealand medical professionals are aware of best practice methods by asking them.

@@

These methods use StatsNZ annual counts of resident deaths and resident population, rather than the weekly counts from mortality.org, as mortality.org takes the raw StatsNZ data and interpolates ages to the weekly totals based on the typical annual age distribution of deaths. This is why New Zealand data for weekly death counts at mortality.org contains decimal places.

Using detailed annual StatsNZ age counts of resident deaths and population prevents inaccuracy in results from overly broad age groups. In 2019 the age specific death rate[^6] of 10-14 year olds (0.17 per 1000 residents of that age per year) approximately doubles by age 15-19 (0.41), and doubles again by 35-39 (0.83). More doubling takes place by 45-49 (1.95), 55-59 (4.45), and 65-69 (10.27). With the speed of doubling being even faster in older aged that make up the majority of deaths, mortality measurements can be very sensitive to even a few years of change in the population balance.

@@

Karlinsky & Kobak also appears superior for comparisons between countries, as Gibson's results are a worse approximation with ASMR based methods calculated by other countries statistical authorities. The Australian Bureau of Statistic's age based mortality analysis has excess mortality of 3.4% across 2020-2022 [@australianbureauofstatistics2023]. Karlinsky & Kobak's approximation of 4% is much closer to the ABS result than Gibson's result of -3%.

@@

>>>>

For hundreds of years, correctly measuring mortality has been a concern of actuaries, demographers, and health researchers. Accurate measurement should form the basis of insurance risk pricing and government policy. Writing in this journal, Gibson [@gibson2024] asserts that New Zealand authorities' measurement of mortality in the pandemic period is based on a flawed metric. Gibson's own metric, based on regression of deaths with an additive component of total population, is asserted as a more accurate model and produces approximately 4% excess mortality for the 2020-2022 period. If New Zealand authorities are using flawed metrics with significantly different results to reality, this would have profound governance and policy implications.

@@

For centuries, it has been established professional practice to measure mortality using age-adjusted death rates where such data is available [@price1773]. This method supersedes the use of Crude Mortality Rates[^2] by acknowledging that different ages have different risks of death, so comparing age-based risk of death on a uniform basis. In New Zealand, tutorials on making mortality comparisons by using death rates by age and standardising to a reference population have been published since the 1890s [@didsbury1891].

@@

Age Standardised Mortality directly adjusts for both population growth and aging, as it is based on the number of deaths of the current resident population divided by the current resident population, stratified by age group and time period. This directly measures the population change effects Gibson is trying to model, and does so at the level of residents of that age in each time period.

@@

The existence of centuries of mortality measurement establishes that it is best practice in mortality calculations to recognise that people of different ages die at different rates not mentioned in Gibson's article. While New Zealand medical professionals have used the Karlinsky and Kobak method/ Our World in Data graphs in media presentations, it is not established if that was for the designed practical purposes while still being aware of more accurate methods or because K & K was considered definitive. We can check if New Zealand medical professionals are aware of best practice methods by asking them.

@@

For calculating age specific death rates, using detailed annual StatsNZ age counts of resident deaths and population prevents inaccuracy in results from overly broad age groups. In 2019 the age specific death rate[^6] of 10-14 year olds (0.17 per 1000 residents of that age per year) approximately doubles by age 15-19 (0.41), and doubles again by 35-39 (0.83). More doubling takes place by 45-49 (1.95), 55-59 (4.45), and 65-69 (10.27). With the speed of doubling being even faster in older aged that make up the majority of deaths, mortality measurements can be very sensitive to even a few years of change in the population balance. Any method calculating cumulative excess using shorter individual time periods should, if correct, produce the same result for the same total time period.

@@

Karlinsky & Kobak also appears superior for comparisons between countries, as Gibson's results are a worse approximation with ASMR (Age Specific Mortality Rate) based methods calculated by other countries statistical authorities. The Australian Bureau of Statistic's age based mortality analysis has excess mortality of 3.4% across 2020-2022 [@australianbureauofstatistics2023]. Karlinsky & Kobak's approximation of 4% is much closer to the ABS result than Gibson's result of -3%.

@@

From comparison to best practice ASMR methods, K&K are closer in result than Gibson's method. Further, given the results of ASMR, New Zealand and international professionals have a sound basis for determining New Zealand's excess mortality as negative for the covid period.
